---
title: "EDA_Project"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Candidate Contributions by Alejandro Garcia
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(tidyr)
library(dplyr)
library(data.table)

options(scipen = 5)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
fc <- read.csv('P00000001-TX.csv', row.names = NULL)
#row.names shifts headers, this fixes it
colnames(fc) <- c(colnames(fc)[-1],"x") 
fc$x <- NULL

```

The following dataset is about the Financial contributions that took place 
for the state of Texas during the 2016 election year. It contains data on 
minor information on the contributors and the amounts they contributed to the 
candidates as well as candidate names, commitee ids, and candidate ids.

________________________________________________________________________________

# Univariate Plots Section

```{r echo=FALSE}
#Structure 
str(fc)
```

Structure of the dataset

```{r echo=FALSE}
#Dimension
dim(fc)
```

Dataset dimensions

```{r echo=FALSE}
head(fc)
tail(fc)
# Gathers summary data on each cand_nm and gives contribution ranges per cand_nm
with(fc, by(contb_receipt_amt,cand_nm , summary))
```

Head and tail data and contribution candidate summary

```{r echo=FALSE, Univariate_Plot_Contribution_per_cand_count}
qplot(data = fc, x = cand_nm, color= I('black')) +
  theme(axis.text.x = element_text(angle = 90))
```

Contributions made to each candidate

```{r echo=FALSE}
#contribution amount brackets
fc$contb_amt.bracket <- cut(fc$contb_receipt_amt, 
                            breaks = c(-16601,0,200,500,1000,2000,15000), 
                            labels = c('<0','0-200','201-500','501-1000',
                                       '1001-2000','2001 >'))
table(fc$contb_amt.bracket)
```

Contribution brackets

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_Brackets_Plot}
ggplot(fc, aes(contb_amt.bracket)) + geom_histogram(stat = 'count')
```

A plot of the contribution amount brackets

```{r echo=FALSE, Univariate_Plot_Whole_view_and_close_up}
qplot(data = fc, x = contb_receipt_amt, color= I('black'),binwidth = 10) +
  coord_cartesian(xlim = c(-1000, 1000)) +
  scale_x_continuous(breaks = seq(-1000,1000,50)) +
  theme(axis.text.x = element_text(angle = 45))

qplot(data = fc, x = contb_receipt_amt, color= I('black'),binwidth = 10) +
  coord_cartesian(xlim = c(-200, 200)) +
  scale_x_continuous(breaks = seq(-200,200,50)) +
  theme(axis.text.x = element_text(angle = 45))

summary(fc$contb_receipt_amt)
```

Contribution Amount Summary and Charts

```{r echo=FALSE}
# Separate contb_receipt_dt into day, month, year
fc<-separate(fc, "contb_receipt_dt", c("cont_day", "cont_month", "cont_year"), 
             sep = "-")
# Order months
fc$cont_month<- ordered(fc$cont_month, levels= c('JAN', 'FEB','MAR','APR','MAY'
                                                 ,'JUN','JUL','AUG','SEP','OCT'
                                                 ,'NOV','DEC'))

# Look at day,month, year frequency
table(fc$cont_day)
table(fc$cont_month)
table(fc$cont_year)
```

Creation of contribution day, month, and year columns.

```{r echo=FALSE, Univariate_Plot_Contribution_Days_Month_and_Year_view}
qplot(data = fc, x = cont_day) +
  facet_wrap(~cont_month, ncol=3,scales = 'free') +
  theme(axis.text.x = element_text(angle = 90))
```

Contributions made per day split by month and containing all years.

# Univariate Analysis

### What is the structure of your dataset?
The current structure of my dataset consist of 548,396 
observations and 21 variables.

### What is/are the main feature(s) of interest in your dataset?
I am interested in finding the amount of contributions made to those 
candidates running for the 2016 presidential elections in the state of TX.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
cand_nam, contbr_nm, contb_receipt_amt, cont_day,month, and 
contb_amt.bracket

### Did you create any new variables from existing variables in the dataset?
I had to create the cont_day, cont_month_, and cont_year which replaced the 
contbr_receipt_dt. The contb_amt.bracket was also created to help reduce the 
amount of bins needed to capture all of the dollar amounts contributed.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Of the features I investigated, I did not notice any unusual distributions. But
I did notice that contbr_zip and contbr_city were not usable since some zip 
codesand cities were made up or other countries.

________________________________________________________________________________

# Bivariate Plots Section



```{r echo=FALSE, Bivariate_Plots_Boxplot_zoom_in}


ggplot(fc, aes(cand_nm, contb_receipt_amt)) + 
  geom_boxplot()+ scale_y_continuous(breaks = seq(-16800,16000,2000)) +
  theme(axis.text.x = element_text(angle = 90)) # Full veiw 

ggplot(fc, aes(cand_nm, contb_receipt_amt)) + 
  geom_boxplot()+
  scale_y_continuous(breaks = seq(-16800,16000,2000)) +
  coord_cartesian(ylim = c(-5000,5000)) +
  theme(axis.text.x = element_text(angle = 90)) #limited veiw

ggplot(fc, aes(cand_nm, contb_receipt_amt)) + 
  geom_boxplot()+
   scale_y_continuous(breaks = seq(-16800,16000,200)) +
  coord_cartesian(ylim = c(-200,3000)) +
  theme(axis.text.x = element_text(angle = 90)) # capture quartiles
```

Zooming in on candidate contribution amounts

```{r echo=FALSE}
mean(fc$contb_receipt_amt)
median(fc$contb_receipt_amt)
summary(fc$contb_receipt_amt)
```

Contribution amount mean, median, and summary

```{r echo=FALSE}
sort(by(fc$contb_receipt_amt, fc$cand_nm, sum),decreasing = T)
sort(by(fc$contbr_nm, fc$cand_nm, length), decreasing = T)
```

Contribution amounts by candidate and contributors count by candidate.

```{r echo=FALSE}
#Create the fc.sum data frame to capture the sum of contribution amounts and the
# total number of contributors each candidate had....not unique, but total count
fc.sum <-  fc %>% 
  group_by(cand_nm) %>% 
  summarise(
    total_contb_amt = sum(contb_receipt_amt),
    total_contb_ct = length(contbr_nm))
# Sorts the total_contb_amt column from highest to lowest amount
fc.sum$cand_nm <- factor(fc.sum$cand_nm, 
                      levels = fc.sum$cand_nm[order(-fc.sum$total_contb_amt)])

fc.sum
```

fc.sum Data all rows

```{r echo=FALSE, Bivariate_Plots_Barplot_cand_contribution_and_contributors_ct}
ggplot(aes(y=total_contb_amt, x=cand_nm), data=fc.sum) + 
  geom_bar(stat= 'identity') + 
  scale_y_continuous(breaks = seq(0,23e+7,2e+6)) +
  theme(axis.text.x = element_text(angle = 90))

ggplot(aes(y=total_contb_ct, x=cand_nm), data=fc.sum) + 
  geom_bar(stat= 'identity') + 
  theme(axis.text.x = element_text(angle = 90))

```

The barplot shows total contribution amounts in the first plot and total 
contributors in the second plot

```{r echo=FALSE, Bivariate_Plots_Total_contributions_by_month_all_years}
ggplot(data = fc, aes(x=cont_day,y=sum(contb_receipt_amt))) + 
  geom_bar(stat='identity')+ 
  facet_wrap(~cont_month,scales = 'free_y')  + 
  theme(axis.text.x = element_text(angle = 90))# sum of contb amt per day
```

Sum of contributions made each day by month and combined years

```{r echo=FALSE, Bivariate_Plots_Number_contrib_made_on_each_day_combined}
ggplot(fc, aes(cont_day,length(contbr_nm))) + 
  geom_bar(stat = 'identity')
```

Number of contributions made on each day for all combined years


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

In this section I wanted to look at the amount of money a candadite received
everyday for the two or three years where contributors donated money. I also 
focused on the combined number of contributions in a month.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

In this section I decided to create a new set of data that contained the 
contribution amounts each candidated received and the amount of contributors.
Althought Hillary Clintion did not win the election she had the highest amount
of contributors with 203K and the highest contribution amount of $22.6 million.

### What was the strongest relationship you found?

Unfortuntaly I was unable to perform a cor test to find any strong correlations 
between the data my dataset contained. The majority of my data is factor based
and numerical values are needed to perform a cor test.

________________________________________________________________________________

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_95_of_contb}
qplot(data = fc, x = contb_receipt_amt, 
      fill= contb_amt.bracket,
      color = I('black'),
      binwidth = 10) +
  xlim(0, quantile(fc$contb_receipt_amt, 0.95)) +
  coord_cartesian(xlim = c(0, 500)) 

#of the contributions submitted, 95% of the amounts contributed are under $500
```

Plot showing 95% of the contributions made to all candidates

```{r echo=FALSE, Contributions_less_than_or_equalto_500_dollars}
with(subset(fc, contb_receipt_amt <=500 ),sum(contb_receipt_amt))
```

Sum of contributions less than or equal to $500

```{r echo=FALSE, Contributions_less_than_or_equalto_2000_dollars}
with(subset(fc, contb_receipt_amt <=2000 ),sum(contb_receipt_amt))
```

Sum of contributions less than or equal to $2000

```{r echo=FALSE, Contributions_in_DEC_of_16}
with(subset(fc, cont_month == 'DEC' & cont_year == '16' & contb_receipt_amt >0 ),sum(contb_receipt_amt))
```

Sum of contributions made in December of 2016

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_contb_by_day}

ggplot(fc, aes(cont_day,length(contbr_nm),color=cont_month)) + 
  geom_bar(stat = 'identity')

# number of contributors by day colored by month 
```

Number of contributors by day and colored by month

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_Box_brackets}
ggplot(data =subset(fc, cont_year =='16' & !contb_amt.bracket =='<0'), aes(x=cont_month,y=contb_receipt_amt,color = contb_amt.bracket)) + 
  geom_boxplot() +
  facet_wrap(~cont_year,ncol=1,scales = 'free_y')
  #theme(axis.text.x = element_text(angle = 90))# sum of contb amt per day
```

Contribution amounts divided by months and contribution amount brackets

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_Contb_month}
ggplot(data = fc, aes(x=cont_day,y=sum(contb_receipt_amt), color= cont_year)) + 
  geom_bar(stat='identity')+ 
  facet_wrap(~cont_month,scales = 'free_y')  + 
  theme(axis.text.x = element_text(angle = 90))# sum of contb amt per day
```

Total contributions made during each day, divided by months, and colored by year

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_contrb_cand}
ggplot(aes(y=total_contb_ct, x=cand_nm, fill=total_contb_amt), data=fc.sum) + 
  geom_bar(stat= 'identity') + 
  theme(axis.text.x = element_text(angle = 90))
```

Contributions each candidate received during the 2016 election colored by amount

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The contribution amounts began in 2013 with contribution amount picking up in 
2015 and amounts sky rocketing in 2016. We observed that there were quite a few
contributors in 2015 with the contributions picking up in November and December. 
After the elections were over, a few contributors continued to contribute until
late December 2016, over a month since the elections ended, thats also where we 
can observe the majority of the contributions over $2001 occurring.

### Were there any interesting or surprising interactions between features?

95% of the contribution amounts were under $500 dollars which equate to 
$26,654,520. As mentioned above during the month of December of 2016 
$76994.23 was contributed to the campains.



# Final Plots and Summary

```{r echo=FALSE}
# Creates a new table to help with pulling the dollar value with the highest ct
DT <- data.table(fc$contb_receipt_amt)
#stores sorted DT into a new data frame
new_DT <- sort(table(DT), decreasing = T)

head(new_DT)
```

The 5 highest dollar amount counts contributed to the 2016 election

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plots_95_of_contb_final}
qplot(data = fc, x = contb_receipt_amt, 
   fill= contb_amt.bracket,
   color = I('black'),binwidth = 10) +
   xlab('Dollar Amounts Contributed') +
   ylab('Count') + 
   labs(title='Majority Contribution Amounts') +
   theme(plot.title = element_text(hjust = 0.5)) +
   scale_fill_discrete(name = 'Contribution\nBrackets') +
   xlim(0, quantile(fc$contb_receipt_amt, 0.95))
```

### Description One

Here we take a look at the contribution amounts by brackets. 95% of the 
contribution amounts were under $500. There were over 80,000 contributions of 
$25 made with the next highest contribution amount being $50 with a count of 
69,794.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plots_Contb_month_final}
ggplot(data = fc, aes(x=cont_day,y=sum(contb_receipt_amt), color= cont_year)) + 
  geom_bar(stat='identity')+ 
  xlab('Day of Contribution') +
  ylab('Contribution Amounts') + 
  labs(title='Contribution by Day and Month') +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(name = 'Contribution Years') +
  facet_wrap(~cont_month,scales = 'free_y')  + 
  theme(axis.text.x = element_text(angle = 90))
```

### Description Two

This shows us a greater picture showing the total amount of money contributed
per day, by month, and year. We can largly see that 2016 had the highest number
of monetary contributions followed by 2015. The amounts donated during 2013 and 
2014 are almost nonexistent on this chart.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plots_contb_by_day_final}
ggplot(fc, aes(cont_day,length(contbr_nm),color=cont_month)) + 
  geom_bar(stat = 'identity') +
  xlab('Contribution by Day') + 
  ylab('Count of Contributors') +
  labs(title='Total Amount of Contributors on a Given Day') +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(name = 'Contribution\nMonth')
```

### Description Three

This shows the total number of contributions made on each day in a four year 
period and colored by month. Each color block represents the number of 
contributions made on that day.

--------------------------------------------------------------------------------

# Reflection

So this was a pretting interesting data set to work with. It contained the 
presidential election contribution information on the candidates and those
who contributed to the campaigns. The dataset was based on the contributions 
from the great state of Texas. Although the state of Texas itself is a mostly
Republican state there was quite a surprise when I found that a Democrat won
out on the number of contributions made and the total amount of contributions
made. I did have a bit of trouble with the dataset itself though, it was mostly
comprived of strings and factors unlike the daimonds dataset which makes it 
easier to work with since it has a ton of numeric based data. The main variables
used were cand_nm, contb_nm, contb_receipt_amt, cont_day, cont_month,cont_year,
and contb_amt.bracket. If I were to work with this dataset again in the future
I might try and divide candidates into their respective political groups.

